Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-29 Manuscript/manuscript.Rmd
edits

[include number of cell type specific features and number of features in common]

To establish the feature space used for classification, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classifier performance in class imbalance settings.

@@

![**Ensemble Framework.** A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline. TAD boundary data was extracted from a specific Hi-C resolution and a variety of functional genomic annotations were used to establish the feature space. Each subsequent cell line specific random forest classification model was built on an isolated predictor type in conjunction with a particular data re-sampling technique to account for class imbalance. The model was then validated on a testing set and performances were evaluated using MCC.](Figures/figure3_ensemble_framework_V2.png)

## CLASS IMBALANCE SEVERELY IMPACTS MODEL PERFORMANCE

We first evaluated the least granular random forest classifier, that was one implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 4A (red)). We then assessing models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no clear improvement (Figure 4B (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance.

>>>>

A total of 107 and 152 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line respectively. To establish the feature space used for classification, we constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classifier performance in class imbalance settings.

@@

![**Ensemble Framework.** A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline. Cell line specific TAD boundary data was extracted from a Hi-C produced contact matrix of a given resolution. A variety of functional genomic annotations were used to establish the feature space. Each subsequent random forest classification model was trained on an isolated predictor type in conjunction with a particular data re-sampling technique to account for class imbalance. Each model was then validated on a testing set and performances were evaluated using MCC.](Figures/figure3_ensemble_framework_V3.png)

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE FOR CELL LINE SPECIFIC MODELS

We first evaluated the least granular cell line specific random forest classifiers, that was one implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 4A (red)). We then assessing models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no clear improvement (Figure 4B (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance.
